A formal approach but practical approach in train regulation

### 2025-06-10 paper/main.tex
Todos

\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{wrapfig}
\usepackage{package/mathpartir}
\usepackage{todonotes}

\usepackage{package/regles}
\usepackage{mathtools}

\newcommand{\directions}{{\mathsf{D}}}
\newcommand{\turnouts}{{\mathsf{T}}}
\newcommand{\blocks}{{\mathsf{B}}}
\newcommand{\internalState}{{\mathcal{I}}}
\newcommand{\forward}{{\mathtt{f}}}
\newcommand{\backward}{{\mathtt{b}}}
\newcommand{\sucblock}{{\mathtt{suc}}}

@@

Final words on the N-scale model. Our circuit has 8 blocks, 5 switches. Each block is bidirectional with a signal at each end. Blocks size goes from 25cm to 70cm.  We operate  three trains in parallel. Due to the size of the model, we reduced our signals  to "red" and "green" status, the trains being too close one from the other. Furthermore, in case of a red signal, we make the hypothesis that the train will be able to stop before it reaches the end of the block. In practice, we stop the trains in the middle.

Routes are not computed in advance. The schedule is renewed every "day"\footnote{We use a time scale factor: one of our "day" corresponds to 20 minutes in real time.}  and we introduce some randomness in the process: beyond some routine daily planning, some trains are inserted/deleted, some trains departure/destination may change, and so on. We simulate also delays. Finally, in our model, we distinguish freight and  passengers routes giving priority to the latter. Routes are recomputed each time a train has finished its duty. The general motivation of this paper is to propose a lightweight scheduling solution that is adapted to the fully automatic control, that takes into account the distributed nature of the system, that is reliable and does not suffer from interlocking.

@@

 Parler de la stratégie, du modèle générale. All these operations are done by what we call  \emph{orders}. Roughly speaking, trains Orders have the shape "Start in direction Until position". Regulator orders are: XXX.

@@

 **** Comparaison avec les travaux antérieurs ****

@@

@@

We suppose that the train network can be described as follows. First, we consider given a set of blocks: $\blocks = \{ 1, 2, \ldots, n\}$. Blocks are oriented: there is a $\forward$ end and a $\backward$ one. Let $\directions = \{\forward, \backward\}$ be the two directions.  Next, we suppose given a set $\turnouts = \{ n+1, \ldots, m\}$  of trackworks (switches, crossings, double slip switches, and so on). Trackworks are characterized by an internal state: switches can be in direct or deviate position (next denoted 'd' and 'v'),  a crossing has only one internal position, a double slip switch can be straight through or crossing. Let $\internalState$ denotes the set of internal states. A \emph{network configuration} is given by a function $\sigma: \turnouts \to \internalState$. Let $\Sigma$ denotes the set of all network configurations.

We end with the function $\sucblock: \blocks \times \directions \times \Sigma \to \blocks$. The functions outputs the first block met according to the starting position, the direction and the network configuration.  From an abstract point of view, the topology of the network reduces to that data.

@@

\subsection{Modélisation d'une liste d'odre}
La modélisation de notre liste d'ordre doit prendre en compte les besoins et contraintes évoqués plus haut.
Pour cela, nous avons choisi de la représenter comme cela : $O = \{\Gamma, R\}$, avec $\Gamma$ l'ensemble des trains et $R$ notre régulateur.
Chaque train peut être représenté par un identifiant unique, une position, une direction et son programme : $t = (id, pos, dir, P)$, avec $t \in \Gamma$.
Le régulateur, portant une grande partie de la logique du modèle, se représente de manière plus complexe : $R = (E,T,S,W,G,H,TL,B)$.
%##########
%Définit ce qu'est un event à ce moment ou plus haut ?
%##########
Nous ne détaillerons pas l'ensemble des éléments de $R$ ici, mais nous pouvons en donner un aperçu rapide : $E$ contient l'ensemble des events,
$T$ le token courant pour chaque ressource et $H$ la dernière position connue de chaque train par le régulateur.

@@

>>>>

\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{wrapfig}
\usepackage{package/mathpartir}
\usepackage{todonotes}
\usepackage{xspace}

\usepackage{package/regles}
\usepackage{mathtools}
\usepackage{ amssymb }

\usepackage{package/formal-grammar}
\usepackage{package/mathpartir}

\newcommand{\Nat}{{\mbox{$\mathbb{N}$}}}

\newcommand{\directions}{{\mathsf{D}}}
\newcommand{\turnouts}{{\mathsf{\theta}}}
\newcommand{\blocks}{{\mathsf{B}}}
\newcommand{\internalState}{{\mathcal{I}}}
\newcommand{\forward}{{\mathtt{f}}}
\newcommand{\backward}{{\mathtt{b}}}
\newcommand{\idle}{{\mathtt{*}}}
\newcommand{\sucblock}{{\mathtt{suc}}}

%% tt style
\newcommand{\sigred}{{\mbox{$\mathtt{r}$}}\xspace}
\newcommand{\siggreen}{{\mbox{$\mathtt{g}$}}\xspace}

\newcommand{\deviate}{{\mbox{$\mathtt{v}$}}\xspace}
\newcommand{\direct}{{\mbox{$\mathtt{d}$}}\xspace}
\newcommand{\incr}{{\mbox{$\mathtt{incr}$}}\xspace}
\newcommand{\turn}{{\mbox{$\mathtt{turn}$}}\xspace}

\newcommand{\su}{{\mbox{$\mathtt{startuntil}$}}\xspace}

\newcommand{\trains}{{\mathsf{T}}}

@@

\paragraph{Final words on the N-scale model.} Our circuit has 8 blocks, 5 switches. Each block is bidirectional with a signal at each end. Blocks size goes from 25cm to 70cm.  We operate  three trains in parallel. Due to the size of the model, we reduced our signals  to "red" and "green" status, the trains being too close one from the other. Furthermore, in case of a red signal, we make the hypothesis that the train will be able to stop before it reaches the end of the block. In practice, we stop the trains in the middle.

\paragraph{Routing trains.}
\todo{Ajouter: tout est automatique}
Routes are computed on the fly. The schedule is renewed every "day"\footnote{We use a time scale factor: one of our "day" corresponds to 20 minutes in real time.}  and we introduce some randomness in the process: beyond some routine daily planning, some trains are inserted/deleted, some trains departure/destination may change, and so on. We simulate also delays. Finally, in our model, we distinguish freight and  passengers routes giving priority to the latter. Routes are recomputed each time a train has finished its duty. The general motivation of this paper is to propose a lightweight scheduling solution that is adapted to the fully automatic control, that takes into account the distributed nature of the system, that is reliable and does not suffer from interlocking.

@@

 Parler de la stratégie, du modèle générale. All these operations are done by what we call  \emph{orders}. Roughly speaking, trains Orders have the shape "Start in direction Until position". Regulator orders are: \todo{en parler}.

@@

\todo{ Comparaison avec les travaux antérieurs / lit review}

@@

\subsection{Notations}

Given some set $S$ and some $x \not\in S$, we denote by $S_x = S \cup \{ x\}$.  An other notation: given some function $f: X \to Y$, $x \in X$ and $y \in Y$, we denote by $f[x \leftarrow y]$, the function:
 $$f[x \leftarrow y] \triangleq \left\{  \begin{array}{ll} x &\mapsto y\\
x' \neq x \in X &\mapsto f(x')
\end{array}\right.$$

@@

We suppose that the train network can be described as follows. First, we consider given a set of blocks: $\blocks = \{ 1, 2, \ldots, n\}$. Blocks are oriented: there is a $\forward$ (forward) end and a $\backward$ (backward) one. Let $\directions = \{\forward, \backward\}$ be the two directions.  Next, we suppose given a set $\turnouts = \{ n+1, \ldots, m\}$, the identifiers  of trackworks (switches, crossings, double slip switches, and so on). Trackworks are characterized by an internal state: switches can be in direct or deviate position (next denoted '\direct' and '\deviate'),  a crossing has only one internal position, a double slip switch can be straight through or crossing. In the examples, to simplify the presentation, we use only switches, but our model is not restricted to them.  Let $\internalState$ denotes the set of internal states. A \emph{network configuration} is given by a function $\sigma: \turnouts \to \internalState$. Let $\Sigma$ denotes the set of all network configurations. For two turnouts $\theta = \{9, 10\}$, the function $9 \mapsto \deviate, 10 \mapsto \direct$ states that $9$ is in deviate position while 10 is in direct one. The latter function is also described via an array  $[\deviate, \direct]$ giving states in trackwork order.

The network is then described as a triple $(\blocks, \turnouts, \sucblock)$ with $\sucblock: \blocks \times \directions \times \Sigma \to \blocks_\bot$. The successor function outputs the first block met according to the starting position, the direction and the network configuration.  The value $\sucblock(pid, d, \sigma) = \bot$ means that one reaches a dead-end. From an abstract point of view, the topology of the network reduces to that data.

On the tracks, we suppose there are sensors to detect the presence of trains. More precisely, we suppose that---by means of the sensors---one is able to compute at any time the position of each train. A \emph{position event} is when a train move from one block to an other one.

Finally, aside of the tracks, at each end of the blocks, there is a signal whose state is \sigred (red) or \siggreen (green). In other words, the status of signals is described by a function $F: \blocks \times \directions \to \{ \sigred, \siggreen\}$. Given some signal status $F$, a block $pid$ and a direction $d$, $F(pid,d) = \sigred$ means a train in block $pid$ cannot leave it in direction $d$. The signal status is entirely determined by the network configuration and the locking status due to the regulator. We will be back to it later.  Since we do not model train speed, we do not distinguish signals with speed limits from full-speed ones.

Now, we come to the train model. Each train has an identifier. Without loss of generality, the identifiers form a set  $\trains = \{1, \ldots, k\}$. Each train has a position on the network. The position is the index of the block it stays in or the last block it left. Next, a train has a direction in $\{\forward, \backward, \idle\}$ with $\idle$ meaning idle status.  The direction does not mean the train is running---it can be blocked by some red signal---but that its next move will be in that direction. Finally, a train has orders. Orders are given as a sequence of orders following the grammar:
$$  seq ::= \su(d,pid) $$
with $d \in \directions$ and $pid \in \blocks$. An order $\su(d, pid)$ states that the train direction must be set to $d$ and that the train runs until it reaches $pid$.  The sequence may be empty.

To sum up, trains can be viewed as 4-tuples: $t = (tid, pid, d, P)$ with $tid$ the train's identifier, $pid$ its position, $d$ its direction and $P$ the sequence of orders. The description of all trains is then a sequence $\Gamma = (t_1, \ldots, t_k)$ with $t_1$ the 4-tuple corresponding to train 1, up to $t_k$ that is the 4-tuple corresponding to train $k$.

To help the reader, by convention, we use $pid$ for blocks, $sid$ for trackworks identifiers (switch id) and $tid$ for train identifiers.

The regulator is in charge of the execution of the planning. For that sake, we use tokens. For each critical resource (either a block or a switch), we associate a token level with each train involved.

More technically, the regulator is characterized by 3 fields. First, it has a general program, next mentioned as $E$. The program is the plan of the regulator. The regulator will follow its rules. A program is a sequence of statements, each statement being composed of a guard and some actions. The guard has the shape: $ev(tid, n)$ meaning that train $tid$ just entered in the $n$-th block of its route. In case such an event occurs, the associated actions are performed. Here is the grammar of the regulator program:
%All statements are guarded by a condition $ev(tid,n)$ stating that train $tid$ reached the $n$-th step of its route. The atoms update the regulator fields. 
%A program is given as a sequence of statements. Those follow the grammar:

\begin{grammar}[Regulator's rules][h][regulateur]
    \firstcasesubtil{stmt}{ev(tid, n) : (atom)*}{}
    \firstcasesubtil{atom}{\incr(pid)}{}
    \otherform{att(pid, v)}{}
    \otherform{\turn(sid, s)}{}
    \otherform{auth}{}
 \end{grammar}
with $tid \in \trains$, $n \in  \Nat$, $pid \in \blocks$, $sid \in \turnouts$, $s\in \internalState$.  
The token field of the regulator, denoted $T$, is a function  $T: \blocks \to \Nat$ associating a token level to each block.  The atom $\incr(pid)$ updates the token $pid$:  $T \cdot \incr(pid)  \triangleq T[pid \leftarrow T(pid)+1]$.

The waiting field is a set $W \subseteq \blocks \times \Nat \times \trains$. A triple $(pid, n, tid) \in W$  meaning intention is that block $pid$ must be protected against train $tid$ entrance as long as $T(pid) < n$.

So, the regulator is described as a tuple $R = (E, T, W)$. Applying $\incr(pid)$ to the regulator is applying it to the token field: $R \cdot \incr(pid) = (E, T \cdot \incr(pid), W)$. We use the same convention for the other atoms.

Finally, the network configuration is given as  $\sigma: \turnouts \to \internalState$. Again, the atom $\turn(sid,s)$ updates the internal state of $sid$: $\sigma \cdot \turn(sid,s) \triangleq \sigma[sid \leftarrow s]$.

%\todo{Je pense que le régulateur ne devrait pas contenir plus d'éléments. Les autres sont en fait des meta-données qui gèrent "le baton de parole" (les priorités sur les règles). Donc, G et B devraient être en dehors du tuple.}

@@

Description par des règles. La gestion du temps est "discrétisée". On ne peut pas "choisir" quelle est la bonne règle. Gestion par des gardes, et des messages. Il y a des soucis grand pas, petit pas. Dire comment on gère les évt, comment on régule l'application des règles.

\subsection{Modélisation d'une liste d'odre}
%La modélisation de notre liste d'ordre doit prendre en compte les besoins et contraintes évoqués plus haut.
%Pour cela, nous avons choisi de la représenter comme cela : $O = \{\Gamma, R\}$, avec $\Gamma$ l'ensemble des trains et $R$ notre régulateur.
%Chaque train peut être représenté par un identifiant unique, une position, une direction et son programme : $t = (id, pos, dir, P)$, avec $t \in \Gamma$.
Le régulateur, portant une grande partie de la logique du modèle, se représente de manière plus complexe : $R = (E,T,S,W,G,H,TL,B)$.
%##########
%Définit ce qu'est un event à ce moment ou plus haut ?
%##########
Nous ne détaillerons pas l'ensemble des éléments de $R$ ici, mais nous pouvons en donner un aperçu rapide : $E$ contient l'ensemble des events,
$T$ le token courant pour chaque ressource et $H$ la dernière position connue de chaque train par le régulateur.

@@

\begin{itemize}
	\item Présenter le logiciel global (TLA + Python + ..., entrées/sorties, etc...)
	\item Schéma architectural
	\item Ordre d'exécution des programmes.
\end{itemize}
